Towards Practical First-Order Model Counting

### 2024-04-12 doc/paper.tex
Writing

% TODO: 9 pages!!!

@@

\title{Towards Efficient First-Order Model Counting}

@@

\begin{abstract}
\end{abstract}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item describe which compilation rules are currently greedy or non-greedy
    \item my agenda has a plan for proving completeness for $\mathbf{C}^{2}$
    \item Can I prove that my domain recursion can always simulate the domain
          recursion of \textsc{ForcLift}?
  \end{itemize}
}

\paragraph{Papers to cite.}
\begin{itemize}
  \item original domain recursion~\cite{DBLP:conf/nips/Broeck11}
  \item recent overview paper~\cite{DBLP:conf/ijcai/Kuzelka23}
  \item conjecturing recurrence relations~\cite{DBLP:conf/ilp/BarvinekB0ZK21}
  \item algorithms
  \begin{itemize}
    \item \textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11}
    \item \textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23}
    \item \textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21}
    \item approximate~\cite{DBLP:conf/ijcai/BremenK20}
  \end{itemize}
  \item liftable fragments (TODO: perhaps good sources of benchmarks?)
  \begin{itemize}
    \item \Ctwo~\cite{DBLP:journals/jair/Kuzelka21}
    \item tree axioms~\cite{DBLP:journals/ai/BremenK23}
    \item linear order axioms~\cite{DBLP:conf/aaai/TothK23}
  \end{itemize}
  \item applications
  \begin{itemize}
    \item extensions to
          sampling~\cite{DBLP:conf/aaai/WangB0K22,DBLP:conf/lics/WangP0K23}
    \item discovery of combinatorial
          sequences~\cite{DBLP:conf/ijcai/SvatosJT0K23}
  \end{itemize}
\end{itemize}

% TODO: make sure to not call \Crane2 \textsc{Crane}

@@

\paragraph{Proof of correctness (leave for later).}
\begin{enumerate}
  \item Define what it means for a formula to be \emph{consistent}.
        \begin{itemize}
          \item The formula (including the mapping of constants to their
                domains) does not mention a domain together with its subdomain
                (i.e., all mentioned domains are pairwise disjoint).
                Specifically, we can't have a subdomain together with its parent
                domain or a subdomain together with a constant that belongs to
                its parent domain.
          \item The target of \texttt{Ref} must also be a consistent formula.
          \item The initial formula is consistent by definition.
        \end{itemize}
  \item Aim to show that \Cranetwo{} works correctly on all consistent formulas.
  \item Prove that, whenever some compilation rule makes the formula $\phi$
        inconsistent, greedy compilation rules will make $\phi$ consistent again
        before $\phi$ encounters non-greedy compilation rules that depend on
        $\phi$ being consistent.
        \begin{itemize}
          \item In particular, prove that shattering and unit propagation will
                always eliminate the parent domain of the domains introduced by
                atom counting.
          \item Constraint removal only applies when it can eliminate the parent
                domain entirely.
          \item Prove this for a particular assumption about which compilation
                rules are set to be greedy. For the greedy algorithm, although
                all rules are applied in a greedy manner, we still assume that
                greedy rules will be applied before non-greedy rules.
        \end{itemize}
  \item In particular, independent partial groundings and atom counting are the
        two classical rules that require the formula to be shattered.
  \end{enumerate}

 % TODO: Crane2 uses itself instead of Crane. Not really. By 'Crane', here I
 % mean the process of turning logical formulas into equations. Later on, I use
 % 'Crane' to refer to Crane2. Maybe rename this use of 'Crane' to something
 % else?
\begin{figure*}[t]
  \centering
  \begin{tikzpicture}
    \node at (0, 0) (formula) {$\phi$};
    \node[draw,rounded rectangle] at (3, 0) (alg1) {\Cref{alg:main}};
    \node[draw,rounded rectangle] at (6, 0) (simplification) {Simplification};
    \node[draw,rounded rectangle] at (9, 0) (compilation) {Compilation};

@@

\paragraph{Contributions (outdated).}
\begin{itemize}
  \item \Cref{sec:identifying}
  \item \Cref{sec:simplifying}
  \item Converting the recursive equations into a C++ program, which can then be
        compiled and executed to obtain numerical values (see \cref{sec:cpp}).
  \item Support for infinite precision integers using the GNU Multiple Precision
        Arithmetic Library.
  \item Some experiments vs FastWFOMC
  \item A proof of completeness
\end{itemize}

@@

@@

\begin{itemize}
  \item Variables in uppercase
  \item Domains in uppercase Greek
  \item Constant and predicate symbols in lowercase
  \item Clauses written on separate lines are implicitly conjoined
  \item For weights, for any predicate symbol $p$, we write
        $w^{+}(p), w^{-}(p) \in \mathbb{R}$ for its positive and negative
        weights, respectively. All weights not explicitly acknowledged are
        assumed to be equal to one.
\end{itemize}

\todo[inline,caption={}]{
  \begin{itemize}
    \item check if I'm still using this notation in all these cases
    \item explain how this works for function calls too
    \item introduce WFOMC as a problem
  \end{itemize}
}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item What notation am I using for FOL? Introduce equality cardinality
          constraints, counting quantifiers, etc.
    \item Need (W)FOMC semantics. What is an interpretation? What is a model?
    \item Should table caption go above or below the table?
    \item Define terms.
    \item Maybe say ``structure'' instead of interpretation?
    \item What's the right table heading capitalization for KR?
    \item Should I say ``sort'' instead of domain everywhere?
  \end{itemize}
}

\paragraph{Three types of logics}
\begin{itemize}
  \item All three logics are function-free.
  \item Domains are always assumed to be finite.
  \item See \cref{tbl:logics} for a detailed comparison. \FO{} is used as the
        input format for \textsc{ForcLift} and its extensions \textsc{Crane} and
        \Cranetwo{}. \Ctwo{} is used in the literature on \textsc{FastWFOMC} and
        related methods (TODO: citations). \UFO{} is the input format supported
        by a privately-obtained version of \textsc{FastWFOMC}. Note that the
        publicly available version does not support any cardinality constraints.
  \item In the case of \textsc{ForcLift} and its extensions, support for a
        formula as valid input does not imply that the algorithm will be able to
        compile the formula into a circuit or graph suitable for lifted model
        counting. However, it is known that \textsc{ForcLift} compilation is
        guaranteed to succeed on any \FO{} formula without constants and with at
        most two variables (TODO: citation).
\end{itemize}

A \emph{formula} is a set of clauses. A \emph{clause} is of the form
$\forall x_{1} \in \Delta_{1}\forall x_{2} \in \Delta_{2}\dots\forall x_{n} \in \Delta_{n} \phi(x_{1}, x_{2}, \dots, x_{n})$,
where $\phi$ is a disjunction of literals that only contain variables
$x_{1}, \dots, x_{n}$ (and any constants). A \emph{literal} is either an atom or
its negation. An \emph{atom} is of the form $P(t_{1}, \dots, t_{m})$, where $P$
is a predicate, and $t_{1}, \dots, t_{m}$ are terms. A \emph{term} is either a
variable or a constant.

\todo[inline]{explain set-based notation for a clause (iterating over all
  literals)}

For any clause or literal $l$, let $\Vars(l)$ denote the set of variables in
$l$.

For any (bound) variable $v$, let $\Dom(v)$ denote the domain over which $v$ is
quantified. For a literal $l$,
$\Doms(l) \coloneqq \{\, \Dom(v) \mid v \in \Vars(l) \,\}$. Similarly, for any
clause $C$, let $\Doms(C) = \{\, \Dom(v) \mid v \in \Vars(C) \,\}$.

@@

NOTE: function calls such as $f(n-m)$ and $f(n-m-1)$ are ignored.

@@

\begin{algorithm}[t]
  \caption{\protect\Propagate{$\phi$, $\Delta$, $n$} TODO: find the source code and check: (1) what we're doing with unit clauses, (2) whether we just randomly pick any literal to negate}\label{alg2}
  \KwIn{formula $\phi$, domain $\Delta$, domain size $n \in \{\, 0, 1 \,\}$}
  \KwOut{formula $\phi'$}
  $\phi' \gets \emptyset$\;
  \uIf{$n = 0$}{
    \ForEach{clause $C \in \phi$}{
      \lIf{$\Delta \not\in \Doms(C)$}{$\phi' \gets \phi' \cup \{\, C \,\}$}
      $C' \gets \{\, l \in C \mid \Delta \not\in \Doms(l) \,\}$\;
      \If{$\Delta \in \Doms(C)$ {\bf and} $C' \ne \emptyset$}{
        $l \gets \text{an arbitrary literal in } C'$\;\label{line:smoothing1}
        $\phi' \gets \phi' \cup \{\, C' \cup \{\, \neg l \,\} \,\} $\;\label{line:smoothing2}
      }
    }
  }
  \Else{
    $c \gets$ a new constant symbol\;
    \ForEach{clause $C \in \phi$}{
      $C' \gets C$\;
      \ForEach{$v \in \Vars(C)$ with $\Dom(v) = \Delta$}{
        $C' \gets C'[c / v]$\;
      }
      $\phi' \gets \phi' \cup \{\, C' \,\}$\;
    }
  }
\end{algorithm}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item Can I prove that the additions to smoothing outlined in the rest of
          this section `do the right thing'?
    \item I could include more info about the smoothing algorithm that didn't
          make it into the previous paper.
  \end{itemize}
}

@@

@@

% TODO
% \item Whenever the set of unit clauses received from the child node contains two
% formulas $\phi(x)$ and $\forall X \in \Delta\text{.
% } X \ne x \Rightarrow \phi(X)$ (i.e., the only difference between the two
% formulas is that one has the constant $x$ whereas the other one has a variable
% $X \ne x$), merge them into $\forall X \in \Delta\text{. }\phi(X)$.

\begin{description}
  \item[\texttt{Ref}:] nothing.
  \item[Constraint removal:] nothing.
  \item[Domain recursion:]
        \begin{enumerate}
          \item Whenever the set of unit clauses of the child node contains two
                formulas $\phi(x)$ and $\forall X \in \Delta\text{.
                } X \ne x \Rightarrow \phi(X)$ (i.e., the only difference
                between the two formulas is that one has the constant $x$
                whereas the other one has a variable $X \ne x$), merge them into
                $\forall X \in \Delta\text{. }\phi(X)$.
          \item Add smoothing nodes below the domain recursion node for the
                difference between the unit clauses assigned to the domain
                recursion node during Stage 1 and the unit clauses of the child
                node post-processed by the step above. For example, if the child
                node `covers' only $p(x)$, then Stage 1 assigns
                $\forall X \in \Delta\text{. }p(X)$ to the domain recursion
                node. The smoothing node below the domain recursion node then
                has the clause $\forall X \in \Delta\text{.
                } X \ne x \Rightarrow p(X)$.
        \end{enumerate}
\end{description}

@@

\section{Experiments}

>>>>

@@

\title{Towards Practical First-Order Model Counting}

@@

\begin{abstract}
  TODO: 9 pages!!!
\end{abstract}

@@

\paragraph{Papers to cite.}
\begin{itemize}
  \item relevant theoretical work~\cite{DBLP:conf/aaai/MalhotraS22} (there is
        more, e.g., the LICS 2018 paper)
  \item original domain recursion~\cite{DBLP:conf/nips/Broeck11}
  \item recent overview paper~\cite{DBLP:conf/ijcai/Kuzelka23}
  \item conjecturing recurrence relations~\cite{DBLP:conf/ilp/BarvinekB0ZK21}
  \item algorithms
  \begin{itemize}
    \item \textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11}
    \item \textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23}
    \item \textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21}
    \item approximate~\cite{DBLP:conf/ijcai/BremenK20}
  \end{itemize}
  \item liftable fragments
  \begin{itemize}
    \item \Ctwo~\cite{DBLP:journals/jair/Kuzelka21}
    \item tree axioms~\cite{DBLP:journals/ai/BremenK23}
    \item linear order axioms~\cite{DBLP:conf/aaai/TothK23}
  \end{itemize}
  \item applications
  \begin{itemize}
    \item extensions to
          sampling~\cite{DBLP:conf/aaai/WangB0K22,DBLP:conf/lics/WangP0K23}
    \item discovery of combinatorial
          sequences~\cite{DBLP:conf/ijcai/SvatosJT0K23}
  \end{itemize}
\end{itemize}

\todo[inline]{Disambiguate between \protect\Crane, \textsc{Crane}, and
  \Cranetwo}

@@

\begin{figure*}[t]
  \centering
  \begin{tikzpicture}
    \node at (0, 0) (formula) {$\phi$};
    \node[draw,rounded rectangle] at (3, 0) (alg1) {\Cref{alg:main}};
    \node[draw,rounded rectangle] at (6, 0) (simplification) {Simplification};
    \node[draw,rounded rectangle] at (9, 0) (compilation) {Compilation};

@@

\paragraph{Contributions (outdated).}
\begin{itemize}
  \item \Cref{sec:identifying}
  \item \Cref{sec:simplifying}
  \item Converting the recursive equations into a C++ program, which can then be
        compiled and executed to obtain numerical values (see \cref{sec:cpp}).
  \item Support for infinite precision integers using the GNU Multiple Precision
        Arithmetic Library.
  \item Some experiments vs FastWFOMC
\end{itemize}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item What notation am I using for FOL? Introduce equality cardinality
          constraints, counting quantifiers, etc.
    \item Maybe say ``structure'' instead of interpretation?
    \item Should I say ``sort'' instead of domain everywhere?
  \end{itemize}
}

\paragraph{Three types of logics}
\begin{itemize}
  \item See \cref{tbl:logics} for a detailed comparison. The notation introduced
        in the table is standard for \Ctwo{}, new for \UFO{}, and redefined to
        be more specific for \FO{}.
  \item All three logics are function-free.
  \item Domains are always assumed to be finite.
  \item
        \begin{itemize}
          \item \FO{} is used as the input format for
                \textsc{ForcLift}\footnote{\url{https://github.com/UCLA-StarAI/Forclift}}~\cite{DBLP:conf/ijcai/BroeckTMDR11}
                and its extensions
                \textsc{Crane}\footnote{\url{https://doi.org/10.5281/zenodo.8004077}}~\cite{DBLP:conf/kr/DilkasB23}
                and \Cranetwo{}.
          \item \Ctwo{} is discussed in the literature on
                \textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21} and related
                methods~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}
          \item \UFO{} is the input format supported by a version of
                \textsc{FastWFOMC} obtained directly from the authors. Note that
                the publicly available
                version\footnote{\url{https://www.comp.nus.edu.sg/~tvanbr/software/fastwfomc.tar.gz}}
                does not support any cardinality constraints.
        \end{itemize}
  \item Note that, in the case of \textsc{ForcLift} and its extensions, support
        for a formula as valid input does not imply that the algorithm will be
        able to compile the formula into a circuit or graph suitable for lifted
        model counting. However, it is known that \textsc{ForcLift} compilation
        is guaranteed to succeed on any \FO{} formula without constants and with
        at most two variables~\cite{DBLP:conf/nips/Broeck11}.
\end{itemize}

\todo[inline]{Explain arity and its notation. Define `ground'?}

A \emph{formula} is a set of clauses. A \emph{clause} is of the form
$\forall x_{1} \in \Delta_{1}\text{. }\forall x_{2} \in \Delta_{2}\dots\text{
}\forall x_{n} \in \Delta_{n}\text{. }\phi(x_{1}, x_{2}, \dots, x_{n})$, where
$\phi$ is a disjunction of literals that only contain variables
$x_{1}, \dots, x_{n}$ (and any constants). A \emph{literal} is either an atom or
its negation. An \emph{atom} is of the form $p(t_{1}, \dots, t_{m})$, where $p$
is a predicate, and $t_{1}, \dots, t_{m}$ are terms. A \emph{term} is either a
variable or a constant. Throughout the paper, we use set-theoretic notation,
interpreting a formula as a set of clauses and a clause as a set of literals.

For any clause or literal $l$, let $\Vars(l)$ denote the set of variables in
$l$. For any (bound) variable $v$, let $\Dom(v)$ denote the domain over which
$v$ is quantified. For a literal $l$,
$\Doms(l) \coloneqq \{\, \Dom(v) \mid v \in \Vars(l) \,\}$. Similarly, for any
clause $C$, let $\Doms(C) = \{\, \Dom(v) \mid v \in \Vars(C) \,\}$.

\begin{definition}[Model]\label{def:model}
  Let $\phi$ be a formula in \FO{}. For every predicate $p/n$ in $\phi$, let
  ${(\Delta_{i}^{p})}_{i=1}^{n}$ be a list of the corresponding domains (not
  necessarily distinct). Let $\sigma$ be a map from the domains of $\phi$ to
  their interpretations as sets such that:
  \begin{itemize}
    \item the sets are pairwise disjoint, and
    \item the constants in $\phi$ are included in the corresponding domains.
  \end{itemize}
  Then an \emph{interpretation} of $\phi$ (with respect to $\sigma$) is a set
  $M$ of ground literals defined by adding either $p(\mathbf{t})$ or
  $\neg p(\mathbf{t})$ for every predicate $p/n$ in $\phi$ and $n$-tuple
  $\mathbf{t} \in \prod_{i=1}^{n} \sigma(\Delta_{i}^{p})$. An interpretation is
  a \emph{model} if it satisfies $\phi$ (see Appendix~A of~Dilkas and
  Belle~\shortcite{DBLP:journals/corr/abs-2306-04189} for more details).
\end{definition}

\begin{definition}[WFOMC]
  Continuing from \cref{def:model}, for every predicate $p/n$ in $\phi$, let
  $w^{+}(p), w^{-}(p) \in \mathbb{R}$ be its (positive and negative)
  \emph{weights}. Unless explicitly specified otherwise, we assume weights to be
  equal to one. The \emph{(symmetric) weighted first-order model count} (WFOMC)
  of $\phi$ (with respect to $\sigma$, $w^{+}$, and $w^{-}$) is the quantity
  \[
    \sum_{M \models \phi} \prod_{p(\mathbf{t}) \in M} w^{+}(p) \prod_{\neg p(\mathbf{t}) \in M} w^{-}(p),
  \]
  where the sum is over all models of $\phi$.
\end{definition}

@@

\begin{itemize}
  \item Variables in uppercase
  \item Domains in uppercase Greek
  \item Constant and predicate symbols in lowercase
  \item Clauses written on separate lines are implicitly conjoined
\end{itemize}

\todo[inline,caption={}]{
  \begin{itemize}
    \item check if I'm still using this notation in all these cases
    \item explain how this works for function calls too
  \end{itemize}
}

@@

@@

@@

\begin{algorithm}[t]
  \caption{\protect\Propagate{$\phi$, $\Delta$, $n$}}\label{alg2}
  \KwIn{formula $\phi$, domain $\Delta$, domain size $n \in \{\, 0, 1 \,\}$}
  \KwOut{formula $\phi'$}
  $\phi' \gets \emptyset$\;
  \uIf{$n = 0$}{
    \ForEach{clause $C \in \phi$}{
      \lIf{$\Delta \not\in \Doms(C)$}{$\phi' \gets \phi' \cup \{\, C \,\}$}
      $C' \gets \{\, l \in C \mid \Delta \not\in \Doms(l) \,\}$\;
      \If{$\Delta \in \Doms(C)$ {\bf and} $C' \ne \emptyset$}{
        $l \gets \text{an arbitrary literal in } C'$\;\label{line:smoothing1}
        $\phi' \gets \phi' \cup \{\, C' \cup \{\, \neg l \,\} \,\} $\;\label{line:smoothing2}
      }
    }
  }
  \Else{
    $c \gets$ a new constant symbol\;
    \ForEach{clause $C \in \phi$}{
      $C' \gets C$\;
      \ForEach{$v \in \Vars(C)$ with $\Dom(v) = \Delta$}{
        $C' \gets C'[c / v]$\;
      }
      $\phi' \gets \phi' \cup \{\, C' \,\}$\;
    }
  }
\end{algorithm}

@@

\todo[inline]{Can I prove that the additions to smoothing outlined in the rest
  of this section `do the right thing'?}

@@

\subsection{Smoothing the FCG}

\begin{algorithm}
  \caption{Propagate atoms for smoothing across the FCG}\label{alg:smoothing}
  \SetKwData{changed}{changed}
  \KwIn{FCG $(V, s, N^+, \tau)$}
  \KwIn{function $\iota$ that maps vertex types in $\mathcal{T}$ to sets of atoms}
  \KwIn{functions $\{\,f_t\,\}_{t \in \mathcal{T}}$ that map a list of sets of atoms to a set of atoms}
  \KwOut{function $S$ that maps vertices in $V$ to sets of atoms}
  $S \gets \{\, v \mapsto \iota(\tau(v)) \mid v \in V \,\}$\;
  $\changed \gets \texttt{true}$\;
  \While{\changed}{
    $\changed \gets \texttt{false}$\;
    \ForEach{vertex $v \in V$}{
      $S' \gets f_{\tau(v)}(\langle S(w) \mid w \in N^+(v) \rangle)$\;
      \If{$S' \ne S(v)$}{
        $\changed \gets \texttt{true}$\;
        $S(v) \gets S'$\;
      }
    }
  }
\end{algorithm}

\todo[inline]{Insert motivation for smoothing from Section 3.4.\ of the ForcLift
  paper.}

Originally, smoothing was (and still is) a two-step process. First, atoms that
are still accounted for in the circuit are propagated upwards. Then, at vertices
of certain types, missing atoms are detected and additional sinks are created to
account for them. If left unchanged, the first step of this process would result
in an infinite loop whenever a cycle is encountered. \Cref{alg:smoothing}
outlines how the first step can be adapted to an arbitrary directed graph.

@@

We never need to add smoothing nodes after \texttt{Ref} or constraint removal
nodes. However, for domain recursion we must do the following.

\begin{enumerate}
  \item Whenever the set of unit clauses of the child node contains two formulas
        $\phi(x)$ and $\forall X \in \Delta\text{.
        } X \ne x \Rightarrow \phi(X)$ (i.e., the only difference between the
        two formulas is that one has the constant $x$ whereas the other one has
        a variable $X \ne x$), merge them into $\forall X \in \Delta\text{.
        }\phi(X)$.
  \item Add smoothing nodes below the domain recursion node for the difference
        between the unit clauses assigned to the domain recursion node during
        Stage 1 and the unit clauses of the child node post-processed by the
        step above. For example, if the child node `covers' only $p(x)$, then
        Stage 1 assigns $\forall X \in \Delta\text{. }p(X)$ to the domain
        recursion node. The smoothing node below the domain recursion node then
        has the clause $\forall X \in \Delta\text{. } X \ne x \Rightarrow p(X)$.
\end{enumerate}

@@

\section{Experimental Evaluation}
